Transactional Memory Concurrency Verification with Landslide

### 2018-03-15 htm/paper.tex
the next step is to colored it.

Our implementation of HTM has been incorporated by the Landslide maintainers upstream.
The repository is available open-source at \url{https://github.com/bblum/landslide}.
Programs should be ported to the Pebbles userland \cite{thrlib,kspec},
their use of compiler HTM intrinsics should be replaced with the Landslide stubs provided in {\tt 410user/inc/htm.h},
and HTM nondeterminism can then be enabled with the {\tt -X} command-line flag.
We have also extended its Iterative Deepening implementation \cite{quicksand}
with an option ({\tt -M}) to
to optimize for completion time by prioritizing the maximal state space job and cancelling all others,
while maintaining its soundness guarantee,
which we will use in our evaluation.
All test cases therein are also available in the repository linked above.
%Several example tests are also provided:
%{\tt htm1} demonstrates the bug in Figure~\ref{fig:htm-example},
%and {\tt htm2} demonstrates Figure~\ref{fig:htm-fixed}'s fix.

@@

\begin{table*}[t]
	\begin{center}
		% TODO: highlight the winning statistic for each test
	\begin{tabular}{cc||r|r|r||r|r|r|r}
			&	&\multicolumn{3}{c||}{\bf Quicksand mode}&\multicolumn{4}{c}{{\bf Maximal state space mode} ({\tt -M})} \\
		buggy test	& params&cputime&walltime&int's&cputime&walltime&int's& SS size (est.) \\
		\hline
		\hline
		{\tt htm1}
			& 2,1	& 45.78	& 9.70	& 21	& 9.47	& 6.40	& 21	& 213 \\
		(assertion)
			& 2,2	& 84.14	& 13.59	& 33	& 10.39	& 7.70	& 49	& 1536 \\
			& 2,3	& 131.91& 20.44	& 73	& 12.83	& 9.67	& 113	& 10752 \\
			& 2,4	& 255.75& 37.56	& 257	& 18.63	& 15.86	& 257	& 73728 \\
		\cline{2-9}
			& 3,1	& 114.06& 17.45	& 15	& 9.50	& 6.79	& 21	& 13653 \\
			& 3,2	& 109.60& 26.16	& 49	& 10.72	& 7.97	& 49	& 393216 \\
			& 3,3	& 124.80& 20.40	& 73	& 13.84	& 11.01	& 113	& 11010048 \\
			& 3,4	& 227.49& 35.15	& 161	& 31.37	& 28.53	& 257	& 301989888 \\
		\cline{2-9}
			& 4,1	& 53.08	& 9.79	& 15	& 9.82	& 7.00	& 21	& 873813 \\
			& 4,2	&117.07	& 19.09	& 33	& 11.54	& 8.55	& 49	& 100663296 \\
		\hline
		{\tt swapbug}
			& 2,1	& 70.95	& 13.45	& 16	& 38.96	& 13.15	& 109	& 194 \\
		(deadlock)
			& 2,2	& 107.28& 17.45	& 146	& 44.73	& 19.47	& 281	& 1620 \\
			& 2,3	& 280.05& 38.70	& 352	& 60.30	& 35.55	& 718	& 12748 \\
			& 2,4	& 617.94& 81.50	& 834	& 108.58& 82.60	& 1820	& 97823 \\
		\cline{2-9}
			& 3,1	&1275.04& 163.42& 771	&{\em >30m}&{\em >30m}&{\em unk.}& 184984 \\
			& 3,2	&{\em >30m}&{\em >30m}&{\em unk.}&{\em >30m}&{\em >30m}&{\em unk.}& 3099225 \\
		% whoops, this was with the dumb lock ordering
		%	& 3,1	& 11.19	& 3.46	& 9	& 3.13	& 3.13	& 9	& 432 \\
		%	& 3,2	& 12.41	& 3.76	& 19	& 3.33	& 3.33	& 19	& 5832 \\
		%	& 3,3	& 15.77	& 4.58	& 41	& 3.95	& 3.95	& 41	& 77759 \\
		%	& 3,4	& 24.82	& 6.85	& 89	& 5.20	& 5.20	& 89	& 1026431 \\
		%\cline{2-9}
		%	& 4,1	& 702.33& 98.06	& 361	& 90.94	& 64.27	& 1607	& 138758 \\
		%	& 4,2	&9738.43&1257.20& 3023	& 674.87& 650.58& 16511	& 13755188 \\
	\end{tabular}
	\end{center}
	\caption{Landslide's bug-finding performance on various test configurations.
		Quicksand's workqueue approach optimized for fast bug-finding
		is compared against our maximum-state-space-prioritizing approach for fast verification.
		For each, we list the CPU-time and wall-clock time (both in seconds)
		elapsed, %until the bug was found,
		plus the number of interleavings of the ultimately buggy state space tested,
		before the bug was found.
		Lastly, Landslide's state space size estimation \cite{estimation},
		though approximate at best,
		confers a sense of the exponential explosion.
	}
	\label{tab:buges}
\end{table*}

@@

Landslide proved the following tests correct using {\tt -M} mode.
With no bugs to find, comparing the different testing modes against each other is meaningless;
we simply present their state space sizes and runtime in Table~\ref{tab:verifs}.
%
Also shown are several
larger tests we were unable to complete before the submission deadline,
each given a timeout of 1 hour. % TODO: ??
Their completion times are listed as {\em unk.}
and their listed state space sizes represent Landslide's estimate at the moment of timeout.
%For these, Table~\ref{tab:timeouts} presents their size estimates at the moment of timeout.
Future work may expand our coverage upon this frontier \cite{landslide-phdthesis}.

\begin{table}[h]
	\begin{center}
		\begin{tabular}{cc|r|r}
			test & params & cputime (s) & SS size \\
			\hline
			{\tt counter}
			& (2,1) & 5.57	& 30	\\
			& (2,2) & 15.53	& 384	\\
			& (2,3) & 20.00	& 5280	\\
			& (2,4) &2211.10& 75264	\\
			& (3,1) & 57.9	& 1960	\\
			& (3,2) &	&	\\
			\hline
			{\tt htm2}
			& (2,1) & 18.57	& 294	\\
			& (2,2) & 133.78& 4902	\\
			& (2,3) &1986.98& 79017	\\
			& (3,1) &11672.15& 467730	\\
			& (3,2) &	&	\\
			\hline
			{\tt swap}
			& (2,1) & 38.77	& 228	\\
			& (2,2) &	&	\\
			\hline
			{\tt map\_basic}
			& (2,1) &	&	\\
			\hline
			{\tt map\_basicer}
			& (2,1) & 877.44& 28635	\\
			& (2,2) &	&	\\
		\end{tabular}
	\end{center}
	\caption{Transactional tests verified by Landslide.}
	\label{tab:verifs}
\end{table}

%\begin{table}[h]
%	\begin{center}
%		\begin{tabular}{cc|r}
%			test & params & SS size (est.) \\
%			\hline
%		\end{tabular}
%	\end{center}
%	\caption{Transactional test verification timeouts.}
%	\label{tab:timeouts}
%\end{table}

>>>>

Our implementation of HTM has been incorporated by the Landslide maintainers upstream.
The repository is available open-source at \url{https://github.com/bblum/landslide}.
Programs should be ported to the Pebbles userland \cite{thrlib,kspec},
their use of compiler HTM intrinsics should be replaced with the Landslide stubs provided in {\tt 410user/inc/htm.h},
and HTM nondeterminism can then be enabled with the {\tt -X} command-line flag.
We have also extended its Iterative Deepening implementation \cite{quicksand}
with an option ({\tt -M} flag) to
to optimize for completion time by prioritizing the maximal state space job and cancelling all others,
while maintaining its soundness guarantee,
which we will use in our evaluation.
All test cases therein are also available in the repository linked above.
%Several example tests are also provided:
%{\tt htm1} demonstrates the bug in Figure~\ref{fig:htm-example},
%and {\tt htm2} demonstrates Figure~\ref{fig:htm-fixed}'s fix.

@@

\definecolor{darkcyan}{RGB}{0,128,128}
\definecolor{lime}{RGB}{48,128,0}
\definecolor{pinkish}{RGB}{128,34,102} % ff4488
\definecolor{brownish}{RGB}{128,96,0}
\newcommand\ETA[1]{\hilight{brownish}{{\em #1}}\xspace}
\newcommand\cpu[1]{\hilight{darkcyan}{{#1}}\xspace}
\newcommand\wtm[1]{\hilight{lime}{{#1}}\xspace}
\newcommand\ints[1]{\hilight{pinkish}{{#1}}\xspace}

\begin{table*}[t]
	\begin{center}
		% TODO: highlight the winning statistic for each test
	\begin{tabular}{cc||r|r|r||r|r|r|r}
			&	&\multicolumn{3}{c||}{\bf Quicksand mode}&\multicolumn{4}{c}{{\bf Maximal state space mode} ({\tt -M})} \\
		\bf buggy test	& \bf params&\cpu{\bf cputime}&\wtm{\bf walltime}&\ints{\bf int's}&\cpu{\bf cputime}&\wtm{\bf walltime}&\ints{\bf int's}& \ETA{\bf SS size (est.)} \\
		\hline
		\hline
		{\tt htm1}
			& 2,1	&\cpu{  45.78}&\wtm{9.70}&\ints{21}& \cpu{9.47}& \wtm{6.40}& \ints{21	}& \ETA{213} \\
		(assertion)
			& 2,2	&\cpu{  84.14}&\wtm{13.59}&\ints{33}& \cpu{10.39}& \wtm{7.70}& \ints{49	}& \ETA{1536} \\
			& 2,3	&\cpu{ 131.91}&\wtm{20.44}&\ints{73}& \cpu{12.83}& \wtm{9.67}& \ints{113	}& \ETA{10752} \\
			& 2,4	&\cpu{ 255.75}&\wtm{37.56}&\ints{257}& \cpu{18.63}& \wtm{15.86}& \ints{257	}& \ETA{73728} \\
		\cline{2-9}
			& 3,1	&\cpu{ 114.06}&\wtm{17.45}&\ints{15}& \cpu{9.50}& \wtm{6.79}& \ints{21	}& \ETA{13653} \\
			& 3,2	&\cpu{ 109.60}&\wtm{26.16}&\ints{49}& \cpu{10.72}& \wtm{7.97}& \ints{49	}& \ETA{393216} \\
			& 3,3	&\cpu{ 124.80}&\wtm{20.40}&\ints{73}& \cpu{13.84}& \wtm{11.01}& \ints{113	}& \ETA{11010048} \\
			& 3,4	&\cpu{ 227.49}&\wtm{35.15}&\ints{161}& \cpu{31.37}& \wtm{28.53}& \ints{257	}& \ETA{301989888} \\
		\cline{2-9}
			& 4,1	&\cpu{ 53.08}&\wtm{9.79}&\ints{15}& \cpu{9.82}& \wtm{7.00}& \ints{21	}& \ETA{873813} \\
			& 4,2	&\cpu{117.07}&\wtm{19.09}&\ints{33}& \cpu{11.54}& \wtm{8.55}& \ints{49	}& \ETA{100663296} \\
		\hline
		{\tt swapbug}
			& 2,1	&\cpu{ 70.95}&\wtm{13.45}&\ints{16}& \cpu{38.96}& \wtm{13.15}& \ints{109	}& \ETA{194} \\
		(deadlock)
			& 2,2	&\cpu{ 107.28}&\wtm{17.45}&\ints{146}& \cpu{44.73}& \wtm{19.47}& \ints{281	}& \ETA{1620} \\
			& 2,3	&\cpu{ 280.05}&\wtm{38.70}&\ints{352}& \cpu{60.30}& \wtm{35.55}& \ints{718	}& \ETA{12748} \\
			& 2,4	&\cpu{ 617.94}&\wtm{81.50}&\ints{834}& \cpu{108.58}& \wtm{82.60}& \ints{1820	}& \ETA{97823} \\
		\cline{2-9}
			& 3,1	&\cpu{1275.04}&\wtm{163.42}&\ints{771}&\ETA{>30m}&\ETA{>30m}&\ETA{unk.}& \ETA{184984} \\
			& 3,2	&\ETA{>30m}&\ETA{>30m}&\ETA{unk.}&\ETA{>30m}&\ETA{>30m}&\ETA{unk.}& \ETA{3099225} \\
		% whoops, this was with the dumb lock ordering
		%	& 3,1	& 11.19	& 3.46	& 9	& 3.13	& 3.13	& 9	& 432 \\
		%	& 3,2	& 12.41	& 3.76	& 19	& 3.33	& 3.33	& 19	& 5832 \\
		%	& 3,3	& 15.77	& 4.58	& 41	& 3.95	& 3.95	& 41	& 77759 \\
		%	& 3,4	& 24.82	& 6.85	& 89	& 5.20	& 5.20	& 89	& 1026431 \\
		%\cline{2-9}
		%	& 4,1	& 702.33& 98.06	& 361	& 90.94	& 64.27	& 1607	& 138758 \\
		%	& 4,2	&9738.43&1257.20& 3023	& 674.87& 650.58& 16511	& 13755188 \\
	\end{tabular}
	\end{center}
	\caption{Landslide's bug-finding performance on various test configurations.
		Quicksand's workqueue approach optimized for fast bug-finding
		is compared against our maximum-state-space-prioritizing approach for fast verification.
		For each, we list the CPU-time and wall-clock time (both in seconds)
		elapsed, %until the bug was found,
		plus the number of interleavings of the ultimately buggy state space tested,
		before the bug was found.
		Lastly, Landslide's state space size estimation \cite{estimation},
		though approximate at best,
		confers a sense of the exponential explosion.
	}
	\label{tab:buges}
\end{table*}

@@

Landslide proved the following tests correct using {\tt -M} mode.
With no bugs to find, comparing the different testing modes against each other is meaningless;
we simply present their state space sizes and runtime in Table~\ref{tab:verifs}.

\newcommand\ETAdag[1]{\ETA{\ensuremath{\dagger}#1}\xspace}
\begin{table}[h]
	\begin{center}
		\begin{tabular}{cc|r|r}
			test & params & cputime (s) & SS size \\
			\hline
			\hline
			{\tt counter}
			& (2,1) & 5.57	& 30	\\
			& (2,2) & 15.53	& 384	\\
			& (2,3) & 20.00	& 5280	\\
			& (2,4) &2211.10& 75264	\\
			& (3,1) & 57.90	& 1960	\\
			& (3,2) &	&	\\
			\hline
			{\tt htm2}
			& (2,1) & 18.57	& 294	\\
			& (2,2) & 133.78& 4902	\\
			& (2,3) &1986.98& 79017	\\
			& (3,1) &11672.15& 467730	\\
			& (3,2) &	&	\\
			\hline
			{\tt swap}
			& (2,1) & 38.77	& 228	\\
			& (2,2) &	&	\\
			\hline
			{\tt map\_basic}
			& (2,1) & \ETAdag{10d 16h} & \ETAdag{16388977} \\
			%[JOB 15] progress: 70330/16388977 brs (0.429130%), ETA 10d 16h 34m 15s (elapsed 43m 21s)
			%[JOB 15] TIMED OUT (0.429130%; ETA 10d 16h 34m 15s)
			\hline
			{\tt map\_basicer}
			& (2,1) & 877.44& 28635	\\
			& (2,2) &	&	\\ % currently running
		\end{tabular}
	\end{center}
	\caption{Transactional tests verified (or not) by Landslide.}
	\label{tab:verifs}
\end{table}

Several larger tests we were unable to complete before the submission deadline are also shown.
%each timing out after 1 hour. % TODO: ??
%Their completion times are listed as {\em unk.}
These are indicated with \ETAdag{}
and their listed ETAs and state space sizes represent Landslide's estimate %at the moment of timeout.
after a timeout of 1 hour
(hopefully enough for data-race PPs to saturate and estimates to stablize).
%For these, Table~\ref{tab:timeouts} presents their size estimates at the moment of timeout.
Future work may expand our coverage upon this frontier \cite{landslide-phdthesis}.
%\begin{table}[h]
%	\begin{center}
%		\begin{tabular}{cc|r}
%			test & params & SS size (est.) \\
%			\hline
%		\end{tabular}
%	\end{center}
%	\caption{Transactional test verification timeouts.}
%	\label{tab:timeouts}
%\end{table}
